---
title: "datapackage R package overview"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
<!-- output: pdf_document -->
vignette: >
  %\VignetteIndexEntry{datapackage R Package Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

## Overview

The *datapackage* R package provides an abstraction for collating heterogeneous collections of data 
objects and metadata into a bundle that can be transported and loaded into a single composite file.  

The methods in this package provide a convenient way to load data from common repositories 
such as DataONE into the R environment, and to document, serialize, and save data from R to 
data repositories. 

## Create a Single Object 
The *datapackage* DataObject class is a wrapper that contains both data and system metadata that describes the data. 
The data can be either R `raw` data or a data file, for example a CSV file. The system metadata includes attributes such as the 
object identifier, type, size, checksum, owner, version relationship to other objects, access rules, and other critical metadata. 

The following example shows how to create a DataObject from a CSV file:

```{r}
library(datapackage)
library(uuid)
csvfile <- system.file("extdata/sample-data.csv", package="datapackage")
myId <- paste("urn:uuid:", UUIDgenerate(), sep="")
myObj <- new("DataObject", id=myId, format="text/csv", filename=csvfile)
```

The DataObject `sciObj` now contains the CSV data as well as the system metadata information. 
The *getData* method can be used to extract the data content of a DataObject. Using the example
DataObject:

```{r}
rawData <- getData(myObj)
```

This raw data can be converted back to CSV format using the R commands:

```{r}
tf <- tempfile(fileext=".csv")
write.csv(rawToChar(rawData), tf, quote=F, row.names=F)
```

To retrieve the identifier associated with a DataObject:

```{r}
id <- getIdentifier(myObj)
```

To retrieve the format type:
```{r}
formatType <- getFormatId(myObj)
```

The entire system metadata information for a DataObject can be access directly from the SystemMetadata object contained in the DataObject:

```{r}
str(myObj@sysmeta)
```

The system metadata contains access policy information for the DataObject that could be used
by a data repository that the object is uploaded to. For example, when a DataObject is
uploaded to a [DataONE Member Nodes](https://www.dataone.org/current-member-nodes), the access policy is applied to the uploaded data
and controls access to the data on the Member Node by DataONE users.

Before the DataObject is uploaed, access can be set so that anyone can read the uploaded data:
```{r}
myObj <- setPublicAccess(myObj)
```

Individual access rules can also be added one at a time:
```{r}
myObj <- addAccessRule(myObj, "uid=smith,ou=Account,dc=example,dc=com", "write")
```

Alternatively, multiple access rules can be added:
```{r}
accessRules <- data.frame(subject=c("uid=jsmith,o=Account,dc=example,dc=com",  
                                    "uid=jadams,o=Account,dc=example,dc=org"), 
                          permission=c("write", "changePermission"))
myObj <- addAccessRule(myObj, accessRules)
str(myObj@sysmeta@accessPolicy)
```

The *dataone* R package can be used to upload or download DataObjecs to a DataONE Member Node.
Please see the web page for the [*dataone*](http://github.com/DataONEorg/rdataone) R package and the 
vignettes for more information:

```{r, eval=FALSE}
library(dataone)
vignette("download-data", package="dataone")
vignette("upload-data", package="dataone")
```

## Create a Collection of Objects
        
A DataPackage is a container for a set of DataObjects. A collection of related DataObjects can be placed in
a DataPackage and actions can be performed on it, such as serializing the entire collection
of objects into a packaging file, or uploading all package member objects to a data repository.

This example creates a DataPackage with one DataObject containing metadata and two others containing science data:

```{r}
metadataFile <- system.file("extdata/sample-eml.xml", package="datapackage")
metadataId <- "metadataId"
metadataObj <- new("DataObject", id=metadataId, format="eml://ecoinformatics.org/eml-2.1.0", file=metadataFile)

csvfile <- system.file("extdata/sample-data.csv", package="datapackage")
sciId <- "sciId1"
sciObj <- new("DataObject", id=sciId, format="text/csv", filename=csvfile)

data <- charToRaw("1,2,3\n4,5,6\n")
sciId2 <- "sciId2"
sciObj2 <- new("DataObject", id=sciId2, data, format="text/csv")
```

The identifier values used in this example are simple and easily recognizable for demonstraction
purposes. A more standard unique identifier can be created with the statement:

```{r}
myid <- paste("urn:uuid:", UUIDgenerate(), sep="")
myid
```

Next a DataPackage object is created and the DataObjects added to it:
```{r}
dp <- new("DataPackage")
addData(dp, metadataObj)
addData(dp, sciObj)
# The second object will be added in the next section 
```

Information can also be extracted from the DataPackage. To show the identifiers of the DataObjects that are in the package:

```{r}
getIdentifiers(dp)
```

To show the size of each DataObject in the package:
```{r}
getSize(dp)
```

To extract the data in a DataObject as raw data:
```{r}
sciObjRaw <- getData(dp, sciId)
```

To extract a package member and create from it a separate DataObject:
```{r}
mySciObj <- getMember(dp, sciId)
```

## Relationships Between DataObjects

Relationships between DataObjects in a DataPackage can be recorded in the DataPackage. For example, 
a typical relationship is that a DataObject containing metadata
can describe, or document, DataObjects containing science data. 

Relationships between DataPackage member 
DataObjects can be recorded with the *insertRelationship* method. This relationship between metadata
and is the default relation recordred by *insertRelationship*, so the relationship type doesn't
need to be specified in this case. For example, with the example DataPackage created above, 
we can add the *documents* relationship:

```{r}
insertRelationship(dp, subjectID=metadataId, objectIDs=sciId)
```

These relationships can be examined with the statements:
```{r}
relations <- getRelationships(dp)
relations[,1:3]
```

(This `documents` relationship is defined by the [Citation Typing Ontology](http://vocab.ox.ac.uk/index) (CITO))

A quick way to add this *documents* relationship is to include the metadata object when a science data
object is added to the package:

```{r}
addData(dp, sciObj2, metadataObj)
```

When a second object is included with the `addData` call, the `insertRelationship` method is
automatically called to establish the *documents* relationship between the two objects, as can
be seen with the updated relationships stored in the package:

```{r}
relations <- getRelationships(dp)
# Print just the first relationship for clarity, without the type information columns
relations[,1:3]
```

Relationships can be fully specified, as shown in the following statement that adds a provenence 
relationship between two objects in the example package:

```{r, eval=F}
insertRelationship(dp, subjectID=sciId2, objectIDs=sciId,
                   predicate="http://www.w3.org/ns/prov#wasDerivedFrom") 
``` 

The relationships contained in a DataPackage conform to the [Resource Description Framework](https://www.w3.org/RDF/) (RDF), which is a [World Wide Web Consortium](https://www.w3.org/)
standard for describing web accessible resources.

## Describing The Contents of a DataPackage 

In order to transport a DataPackage, for example to a data repository, it is necessary that a description of
the contents of the DataPackage is created so that the consumer of the DataPackage can 
determine how to extract, and process the contents.

A DataPackage can produce a standard description of its members and relationships which conforms to the 
the Open Archives Initiative [Object Reuse and Exchange](https://www.openarchives.org/ore/) (OAI-ORE), 
which is a widely used standard to describe aggregations of web accessible resources. This OAI-ORE 
description is refered to as a *resource map*.

The *serializePackage* method will create Resource Description Framework serialization of a resource map,
written to a file in this case, that conforms to the OAI-ORE specification.

To create a resource map for the example DataPackage:

```{r}
tf <- tempfile()
packageId <- paste("urn:uuid:", UUIDgenerate(), sep="")
serializePackage(dp, file=tf, id=packageId)
```
This example writes to a tempfile using the default serialization format of "rdfxml". Also the URLs for each 
package member are prepended with the default value of the DataONE resolve service, which would be the URL
that could be used to access this data object if the package is uploaded to a DataONE member node. 

A different value to be prepended to each identifier can be specified with the *resoveURI* argument. 
To specify that no value be prepended to the identifier URLs, specify a zero-length character:

```{r}
tf <- tempfile()
packageId <- paste("urn:uuid:", UUIDgenerate(), sep="")
serializePackage(dp, file=tf, id=packageId, resolveURI="")
```

It is also possible to create a JSON serialization, if desired:

```{r}
tf <- tempfile()
packageId <- paste("urn:uuid:", UUIDgenerate(), sep="")
serializePackage(dp, file=tf, id=packageId, syntaxName="json", mimeType="application/json", resolveURI="")
```

## Saving DataPackage Contents to a File

The contents of a DataPackage can be saved to a file using the *serializeToBagIt* method. This creates 
a *BagIt* file, a hierarchical file packaging format. 

The created BagIt file contains the data from the DataPackage members as well as an OAI-ORE resource
map that is automatically created by *serializeToBagIt*.
 
The following R command shows how to create the BagIt file for the example DataPackage:

```{r, eval=F}
bagitFilename <- serializeToBagIt(dp)
```

The variable `bagitFilename` contains the file path to the temporary BagIt file.
This file should be copied to another location before quiting or restarting R:

```{r, eval=F}
file.copy(bagitFilename, "~/myPackageFile.zip")
```